---
title: "mutation_count_per_microbiopsy"
author: "Natalia Brzozowska"
date: "2023-10-30"
output: html_document
---

# Load libraries

```{r, message=FALSE}
rm(list=ls())

library(dplyr)
library(data.table)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
library(dndscv)
library(ggplot2)
```

# Load and prepare files

```{r}
wgs_calls <- read.table("data/calls/wgs_calls_v2.csv", sep = ",", stringsAsFactors = F, header = T)

donor_info <- read.table("data/metadata_files/A1AD_patient_info_2019-10-30.csv", sep = ",", stringsAsFactors = F, header = T) |>
  dplyr::rename(internal_id=Donor,
         age=Age)

lcm_database <-  fread("data/metadata_files/lcm_cut_metadata.tsv", sep="\t", header=T, data.table=F) 

exome_calls <- read.table("data/calls/exome_calls_v3.csv", sep = ",", stringsAsFactors = F, header = T)

bed <- read.table("data/reference_files/lcm_exome_callable_bait_set.bed",sep="\t",header=T)

```

# Combine the exome and genome calls

A few samples had both WGS and WXS performed. To avoid discovery bias, only cound the snvs detected by WGS.
```{r}

duplicated_samples <- lcm_database$sampleID[duplicated(lcm_database$sampleID)]

wgs_calls_filt <- wgs_calls %>%
  dplyr::select(mut_id,Chrom,Pos,Ref,Alt,sampleID,Mut_Frags,mutID_sampleID,coverage,vaf,Project)

exome_calls_filt <- exome_calls %>%
  dplyr::select(mut_id,Chrom,Pos,Ref,Alt,sampleID,Mut_Frags,mutID_sampleID,coverage,vaf,Project) %>%
  filter(!sampleID %in% duplicated_samples) # for samples that were both wgs and wex, remove the wgs calls to not be biased by doubled coverage

all_calls <- rbind(wgs_calls_filt,exome_calls_filt) %>%
  distinct(mutID_sampleID, .keep_all = TRUE) %>%
  mutate(donor_id=substr(sampleID,1,7)) %>%
  dplyr::rename(chrom=Chrom,
         pos=Pos,
         ref=Ref,
         mut=Alt)

```

# Subset calls to callable exome range from bed file

bed file was generated with the "define_callable_exome_region.Rmd" file
```{r}

# Create a GRanges object for 'all_calls'
all_calls_gr <- GRanges(seqnames = all_calls$chrom,
                        ranges = IRanges(start = all_calls$pos, end = all_calls$pos),
                        mut_id = all_calls$mut_id,
                        sampleID = all_calls$sampleID)

# Create a GRanges object for 'bed'
bed_gr <- GRanges(seqnames = bed$chrom,
                  ranges = IRanges(start = bed$start, end = bed$end))

# Find the intersection between 'all_calls' and 'bed'
intersection <- subsetByOverlaps(all_calls_gr, bed_gr)

# Extract the subset of 'all_calls' that intersects with 'bed'
filtered_calls <- as.data.frame(intersection) %>%
  dplyr::select(seqnames,start,mut_id,sampleID) %>%
  dplyr::rename(chr=seqnames,
         pos=start) %>%
  mutate(mutID_sampleID = paste0(sampleID,"_",mut_id)) %>%
  left_join(all_calls[,c("ref","mut","mutID_sampleID","Mut_Frags","coverage","vaf","donor_id","Project")],by="mutID_sampleID") %>%
  dplyr::select(chr,pos,ref,mut,sampleID,donor_id,mut_id,Mut_Frags,coverage,vaf,Project)

head(filtered_calls)

```

# plot total number of mutations per exome and coverage per sample

```{r}
exome_size <- sum(bed$length)

samples <- unique(filtered_calls$sampleID)

snv_filtered_calls <- filtered_calls[nchar(filtered_calls$ref)==1 & nchar(filtered_calls$mut)==1,]

burden <- as.data.frame(array(data = NA, dim = c(length(samples),5)))
colnames(burden) <-c("internal_id","paper_donor","donor_age","snv_count","snv_burden")

burden$internal_id <- samples

for(i in 1:nrow(burden)){
  burden$paper_donor[i] <- donor_info$paper_id[which(substr(burden$internal_id[i],1,7) == donor_info$internal_id)]
  burden$donor_age[i] <- as.numeric(paste0(donor_info$age[which(substr(burden$internal_id[i],1,7) == donor_info$internal_id)]))
  burden$snv_count[i] <- length(which(snv_filtered_calls$sampleID == burden$internal_id[i]))
  burden$snv_burden[i] <- (burden$snv_count[i] / exome_size) * 1000000
  burden$mean_vaf[i] <- mean(snv_filtered_calls[snv_filtered_calls$sampleID == burden$internal_id[i],"vaf"])
}

burden_liver <- burden %>%
  mutate(disease = case_when(
    startsWith(paper_donor, "HH") ~ "HH",
    startsWith(paper_donor, "AT") ~ "A1AD",
    TRUE ~ NA_character_
  ))


## remove duplicate exome samples, only keep genomes
lcm_database <- lcm_database[!(lcm_database$platform == "WXS" & lcm_database$sampleID %in% duplicated_samples), ]

all_data_sorted <- burden_liver %>%
  group_by(disease, paper_donor) %>%
  mutate(mx = max(snv_count)) %>%
  arrange(disease, donor_age,mx,snv_count) %>%
  dplyr::select(-mx) %>%
  left_join(lcm_database[,c("sampleID","coverage","platform")] %>% dplyr::rename(internal_id=sampleID),by="internal_id") %>%
  mutate(vafdepth = mean_vaf*coverage)


all_data_sorted$internal_id <- factor(all_data_sorted$internal_id, levels = unique(all_data_sorted$internal_id))
all_data_sorted$paper_donor <- factor(all_data_sorted$paper_donor, levels = unique(all_data_sorted$paper_donor))

# plot total mutation counts
counts_plot <- ggplot(all_data_sorted, aes(x=internal_id, y=snv_count, fill=paper_donor, col=paper_donor)) +
  geom_point() +
  scale_fill_manual(values = rep("#5A9D9E",10)) +
  scale_color_manual(values = rep("#5A9D9E",10)) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  theme(strip.text.x = element_text(angle = 45)) +
  theme(legend.position = "none") +
  facet_grid(~paper_donor, switch = "x", scales = "free_x", space = "free_x")    +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 150)) +
  labs( x="Sample",y = "Mutations per exome",  size = 0.9)  +
  theme(panel.spacing=unit(4,"pt"),  panel.border=element_rect(colour="white", fill=NA))
counts_plot


coverage_plot <- ggplot(all_data_sorted, aes(x = internal_id, y = coverage, fill = factor(platform))) +
  scale_fill_manual(values = c("#62C3A6","#839AC5")) +
  geom_bar(stat = "identity", position = position_dodge(0.9), width = 0.6, alpha = 0.7) +
#  scale_fill_manual(values = rep("#5A9D9E", 10)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        strip.text.x = element_text(size=0),
        legend.position = "none",
        panel.grid = element_blank()) +
  facet_grid(~paper_donor, switch = "x", scales = "free_x", space = "free_x") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, max(all_data_sorted$coverage) + 20)) +
  labs(x = NULL, y = "Coverage", size = 0.9) +
  theme(panel.spacing = unit(4, "pt"), panel.border = element_rect(colour = "white", fill = NA))
coverage_plot

combined_plot <- coverage_plot / counts_plot + 
  plot_layout(heights = c(1, 2))

combined_plot


```

# annotate mutations with dnds

```{r}
covs = "data/reference_files/covariates_20pc_GRCh37-38.epi_strict_outliers.Rdat"
load(covs)
refcds_38 = "data/reference_files/refcds_GRCh38-GencodeV18+Appris.rda"
load(refcds_38)

muts <- filtered_calls %>%
  dplyr::select(sampleID,chr,pos,ref,mut) %>%
  mutate(chr=gsub("chr","",chr))

dndsout <- dndscv(muts, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, outmats = T, refdb = refcds_38, cv = scores, mingenecovs = 0)

annotmuts <- dndsout$annotmuts

annotmuts <- annotmuts %>%
  dplyr::select(sampleID,chr,pos,ref,mut,impact) %>%
  dplyr::mutate(internal_id=substr(sampleID,1,7)) %>%
  dplyr::left_join(donor_info[,c("internal_id","paper_id","Disease","age")]) %>%
  mutate(impact = ifelse(impact == "no-SNV", "Indel", impact)) %>%
  mutate(disease = case_when(
    startsWith(paper_id, "HH") ~ "HH",
    startsWith(paper_id, "AT") ~ "A1AD",
    TRUE ~ NA_character_
  )) %>% arrange(disease, age)


  
```

# plot Fig. 1b - the number of coding mutations within callable exome region per sample

```{r}

# Convert "Impact" to a factor with a specified order for better visualization
annotmuts$impact <- factor(annotmuts$impact, levels = c("Synonymous", "Missense", "Nonsense", "Indel", "Essential_Splice", "Stop_loss"))

observ_counts <- annotmuts %>%
  group_by(sampleID) %>%
  summarize(total_observations = n()) %>%
  dplyr::mutate(donor=substr(sampleID,1,7)) %>%
  arrange(donor,total_observations)

# Reorder the sampleID based on total number of observations
annotmuts$sampleID <- factor(annotmuts$sampleID, levels=unique(observ_counts$sampleID))
annotmuts$internal_id <- factor(annotmuts$internal_id, levels=unique(annotmuts$internal_id))

# Create a stacked bar plot
coding_plot <- ggplot(annotmuts, aes(x = sampleID, fill = impact)) +
  geom_bar() +
  facet_grid(~internal_id, scales = "free_x", space="free") +
  scale_fill_manual(values = c(
    "Synonymous" = "#B1B4B4",
    "Missense" = "#4783B1",
    "Nonsense" = "#CE4427",
    "Indel" = "#79C262",
    "Essential_Splice" = "#A360A7",
    "Stop_loss" = "#FAA32C"
  )) +
  theme_minimal() +
  labs(x = "Sample ID", y = "Number of Mutations", fill = "Impact") +
  theme(legend.position = "right",
        axis.text.x=element_blank(),
        panel.grid.minor = element_blank()) +
  theme(panel.spacing = unit(4, "pt"), panel.border = element_rect(colour = "white", fill = NA))
coding_plot
ggsave(plot = coding_plot, filename = "outputs/coding_mut_counts_per_exome_in_callable_bed.pdf", width = 12, height = 4, useDingbats = F)


# Combine the plots using patchwork
combined_plot_2 <- coverage_plot / coding_plot + 
  plot_layout(heights = c(1, 3))
combined_plot_2
ggsave(plot = combined_plot_2, filename = "outputs/coding_mut_counts_per_exome_in_callable_bed_with_coverage.pdf", width = 15, height = 5, useDingbats = F)


```


# Plot Extended Fig. 2 - signature exposures from exomes

```{r}
library(tidyverse)

exposures <- read.table("data/signatures/exposures_per_lcm_sample_all.txt")
exposures=exposures[,c("SBS1","SBS5","SBS40","SBS12","SBS16","Stan_sig","N2")]
exposures[,"SBS5"]=exposures[,"SBS5"]+exposures[,"SBS40"]
colnames(exposures)[colnames(exposures)=="SBS5"]="SBS5/40"
exposures=exposures[,colnames(exposures)!="SBS40"]

colnames(exposures)[colnames(exposures)=="Stan_sig"] <- "SBSLiver"

exposures_long <- exposures %>%
  rownames_to_column(var = "sampleID") %>%
  pivot_longer(cols = -sampleID, names_to = "signature", values_to = "proportion") %>%
  dplyr::mutate(internal_id=substr(sampleID,1,7)) %>%
  dplyr::left_join(donor_info[,c("internal_id","paper_id","Disease","age")]) %>%
  mutate(disease = case_when(
    startsWith(paper_id, "HH") ~ "HH",
    startsWith(paper_id, "AT") ~ "A1AD",
    TRUE ~ NA_character_
  )) %>% arrange(disease, age)

exposures_long$sampleID <- factor(exposures_long$sampleID, levels=levels(annotmuts$sampleID))
exposures_long$paper_id <- factor(exposures_long$paper_id, levels=unique(exposures_long$paper_id))
exposures_long$internal_id <- factor(exposures_long$internal_id, levels=unique(exposures_long$internal_id))

exposures_plot <- ggplot(exposures_long, aes(x = sampleID, y = proportion, fill = signature)) +
  geom_bar(stat = "identity", position = "fill") + # Use stat="identity" for pre-calculated proportions
  facet_grid(~internal_id, scales = "free_x", space = "free") + # Ensure paper_id or similar grouping variable is available
  scale_fill_manual(values = c(
    "SBS1" = "#619947",
    "SBS5/40" = "#686EA8",
    "SBS12" = "#8BCECD",
    "SBS16" = "#9695C9",
    "SBSLiver" = "#292A53",
    "N2" = "#E0E4ED"
  )) +
  theme_minimal() +
  labs(x = "Sample ID", y = "Signature proportions", fill = "Signature") +
  theme(
    legend.position = "right",
    axis.text.x = element_blank(),
    strip.text.x = element_text(hjust = 0), # Angle facet titles
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.title.x = element_blank() # Ensure x-axis title is removed
  )

exposures_plot

ggsave(plot = exposures_plot, filename = "outputs/signature_exposure_proportions_per_exome_lcm_sample_all.pdf", width = 12, height = 4, useDingbats = F)

coding_plot / exposures_plot

```

